---
title: "STA2060 - Quiz"
output: html_document
date: "2024-01-29"
---

# QUIZ 1: STA2060-A

Name: Chesia Anyika

ID: 665567

## Libraries

```{r}
library(DiagrammeR)
```

## Question 1

Suppose that taxis pick up and deliver passengers in a city which is divided into three zones: A, B and C. Records kept by the drivers show that:

-   Of the passengers picked up in zone A, 70% are taken to a destination in zone A, 20% to zone B, and 10% to zone C.
-   Of the passengers picked up in zone B, 50% go to zone A, 25% to zone B, and 25% to zone C.
-   Of the passengers picked up in zone C, 20% go to zone A, 40% to zone B, and 40% to zone C.

Suppose that at the beginning of the day 60% of the taxis are in zone A, 10% in zone B, and 30% in zone C.

**a)** **Design both the state matrix and transition matrix**

The information above can be summarised in the following tables. For the transition matrix:

|       | A   | B   | C   |
|-------|-----|-----|-----|
| **A** | 70% | 20% | 10% |
| **B** | 50% | 25% | 25% |
| **C** | 20% | 40% | 40% |

And for the state matrix:

| A   | B   | C   |
|-----|-----|-----|
| 60% | 10% | 30% |

I created these matrices in R, and changed the percentages to decimal values for ease of computation.

```{r}
#Transition matrix in percentages
tm <- matrix(c(70, 50, 20, 20, 25, 40, 10, 25, 40), 
             nrow = 3, ncol = 3, 
             dimnames = list(c("A", "B", "C"), c("A", "B", "C")))

#Divide the matrix by 100 to convert values into decimals
tm <- tm/100

tm
```

```{r}
#state matrix in percentages
sm <- matrix(c(60, 10, 30), nrow = 1, ncol = 3,
             dimnames = list(c(" "), c('A', 'B', 'C')))

#divide the state matrix by 100 to convert values into decimals
sm <- sm/100

sm
```

**b) Design the transition diagram**

To create the transition diagram I used the `grViz` function from the `DiagrammeR` library.

```{r}
grViz("
      digraph boxes_and_circles {
      
      #graph statement
      graph [overlap = true, fontsize = 10]
      
      #node statements
      node [shape = circle,
      fixedsize = true, 
      width = 0.9
      fontname = Helvetica]
      A; B; C
      
      
      #edge statements with labels
      A->A [label = \"0.7\"]
      A->B [label = \"0.2\"]
      A->C [label = \"0.1\"]
      B->A [label = \"0.5\"]
      B->B [label = \"0.25\"]
      B->C [label = \"0.25\"]
      C->A [label = \"0.2\"]
      C->B [label = \"0.4\"]
      C->C [label = \"0.4\"]
      }
      
      ")
  
  




```

**c) What is the distribution of taxis in the various zones after all have had one rider?**

Initially, as per the state matrix, 60% of the taxis are in zone A, 10% of the taxis are in zone B and 30% of the taxis are in zone C. To get the distribution of taxis in each zone after all taxis have had one rider, we need to multiply the state matrix by the transition matrix as per the formula:

$$S' = S \times T $$ Where :

**S = State matrix**

**T = Transition matrix**

**S' = New state matrix after the transition**

This is shown below:

```{r}
#multiply state matrix sm byt transition matrix tm
sm2 <- sm %*% tm
sm2
```

The result shows that after all taxis have one rider, the distribution of taxis is:

> 53% in Zone A
>
> 26.5% in Zone B
>
> 20.5% in Zone C.

**d) What is the distribution of taxis in the various zones after all have had two riders?**

To get the distribution of taxis in each zone after all have had two riders, we need to square the transition matrix to get the probability matrix after two transitions, then multiply this by the state matrix as per the formula:

$$
S' = S \times T^n
$$

Where **S**, **T** and **S'** are defined as in part c) above, and **n = the number of transitions**.

This is shown below:

```{r}
#square the transition matrix
tm2 <- tm %*% tm
tm2
```

```{r}
#multiply initial state matrix by squared transition matrix
sm3 <- sm %*% tm2
sm3
```

The result shows that after all the taxis have had two riders, the distribution of taxis is:

> 54.45% in Zone A
>
> 25.425% in Zone B
>
> 20.125% in Zone C

**e) What is the distribution of taxis in the various zones after all have had four riders?**

To get the distribution of taxis in each zone after all have had four riders, we need to get the product of the transition matrix to the fourth power, to get the probability matrix after four riders. We then multiply this by the initial state matrix as per the formula:

$$
S' = S \times T
$$

Wher **S, T, S'** and **n** are defined as in part d) above.

```{r}
#product of the transition matrix to the fourth power
tm4 <- tm %*% tm %*% tm %*% tm
tm4
```

```{r}
#multiply the new transition matrix by the initial state matrix
sm4 <- sm %*% tm4
sm4
```

The result shows that after all the taxis have had four riders, the distribution of taxis is:

> 55.02% in Zone A
>
> 25.24% in Zone B
>
> 19.75% in Zone C
